PCA

### 2023-11-21 DRAFT_PCA.Rmd
implemented Dani's changes to PCA page, minor edits

```{r setup, eval=TRUE, echo=TRUE, message = FALSE, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, cache = TRUE) 
```

@@

On this page we outline and describe many aspects of PCA:

* What is PCA and what are its common uses in population genetics?
* Considerations before sampling:
    + How one chooses to sample from a landscape
    + Sample sizes
* Considerations while processing data (aka 'weird stuff that happens in PCA'): 
    + Batch effects
    + LD and inversion effects
    + Missing data and imputation
    + Hybridization and inbreeding effects
    + Horseshoe and wave patterns
* Comparison to other common approaches to investigating population genetic structure
    + PCA  vs ADMIXTURE/STRUCTURE

@@

## What is PCA and what are its common uses in population genetics?

@@

The eigenvectors are represented as a standard scatter plot, where points are samples/individuals represented on the new system of axes (PCs). Dimension reduction to eigenvectrors (PCs) means that the 1st PC is the linear combination of the original variables that explains the most/greatest amount of variation, while the 2nd PC accounts for the greatest amount of the remaining variation being orthogonal (uncorrelated) to the 1st PC, the 3rd PC accounts for the greatest amount of the remaining variation being orthogonal (uncorrelated) to the 1st and 2nd PC so on and so on. In population genomics, typically the 1st PC will capture broad scale population structure (because broadly distributed populations are usually the most differentiated) while subsequent axes capture more regional population structure (usually less differentiated at smaller geographic scales, see Figure 2A). However, other cool examples using PCA find the greatest amount of variation is between individuals, particularly in the case of inversions, structural variation or regions associated with a specific phenotype or adaption (see Figure 2B).

@@

* [Jolliffe, I. & Cadima, J. (2016) Principal component analysis: a review and recent developments](https://doi.org/10.1098/rsta.2015.0202)
* [Pearson K. (1901) On lines and planes of closest fit to systems of points in space. Philos Mag 2:559–572](https://doi.org/10.1080/14786440109462720)
* [Hotelling H (1936) Relations between two sets of variates. Biometrika 28:321–377](https://doi.org/10.1007/978-1-4612-4380-9_14)
* [Fentaw Abegaz et al. (2019) Principals about principal components in statistical genetics, Briefings in Bioinformatics](https://doi.org/10.1093/bib/bby08)
* [Josse J., Husson F. (2012) Handling missing values in exploratory multivariate data analysis methods. J. Soc. Française Stat., 153, 79–99.](http://www.numdam.org/item/JSFS_2012__153_2_79_0/)

@@

## Considerations before sampling
### How one chooses to sample from a landscape
How samples are selected from a metapopulation also affects the visual pattern in a PCA.

@@

### Sample sizes
Uneven sampling between groups, the number of samples and/or the number of markers/SNPs used in PCA can change the PCA projection space, and thus interpretation of results, relative to the true demographic history of the sampled groups.

@@

Missing data in population genomic analysis is common.  Sometimes it is random, or it may be a feature of the data type (i.e. low coverage NGS, microsatellite null alleles, mutation at RAD cut sites). In general, if missing data reflects a true population signal (e.g. null alleles in one population but not others) then it requires careful consideration during analysis.

@@

Implementations of PCA in genomics typically (a) avoid missing values (i.e pcadapt, plink) usually by first determining the covariance-matrix or genomic relationship matrix (GRM)) between each pair of individuals using the variables (loci) that are available for both samples or by (b) impute missing values. Imputation is the process of replacing a missing value with a numerical value via inference. Imputation of missing data can be done by sophisticated methods such as in genomic association studies, linked haplotypes, and strong genotype-phenotype correlations (see @Yi2022-bi). However, in population genomic studies for species of conservation concern which may have few genomic resources, simple imputations such as mean imputation (replace missing with column mean) are commonly encountered.

Imputation via the mean genotype in the metapopulation.
In this case, missing values for all individuals at a locus are replaced with the mean for that locus (also called a mean-imputed matrix). Importantly, when the mean-imputed matrix is centered and scaled, the original missing become non-informative in PCA (missing values are placed at the origin). This can be problematic if you have a lot of missing data in your dataset, especially if that missing data is biased (not random). @Yi2022-bi demonstrated this effect using different simulated populations (no migration. 5% migration, isolation-by-distance-cline) with 1%, 10% and 20% missing data, either randomly distributed among individuals, or biased among some individuals and some populations. They conducted PCA using the adegenet::glPca() implementation of PCA with default mean imputation and standardization, retaining all PCA axes. They found that individuals biased with missing data would be dragged away from their real population clusters to the origin of PCA plots, making them indistinguishable from known admixed individuals, potentially leading to misinterpreted population structure (Figure 11).

@@

**Effects of different approaches to imputation on PCA.**

The choice of imputation can have drastic effects on inference if there is a lot of missing data. When there is population structure and missing genotypes are imputed with the mean genotype across all populations, the genetic differentiation among populations may be artificially reduced and the metapopulation would appear to be panmictic.  When there is no population structure and missing genotypes are imputed with the mean genotype within each sampling location, the genetic differentiation among populations may be artificially inflated and it could appear that their genetic structure when there is in fact none. As such, it is really important to understand the type of imputation a PCA method might be implementing  and its effect  on population clustering and final inferences.

Since most genomic data has missingness, it may be preferable to implement a method of PCA which explicitly accepts it (i.e pcadapt, PCAngsd).

@@

### Horseshoe and Wave patterns
Horseshoe patterns arise in population genetic data that has isolation by distance structure, by which genetic differentiation among locations increases with the distance between them. The result when this type of data is put into a PCA is that the pattern looks like a horseshoe (Figure 1B from @Morton2017-xy below). In Figure 1B, Sample 0 is most distantly related from Sample 20, but they have the same loading on the PC1 axis. Horseshoe patterns also arise in microbial ecology, where they were originally misinterpreted as having an ecological explanation before they were proven to be a statistical artifact (@Morton2017-xy). The pattern arises in microbial data or genetic data as a consequence of distance metrics that saturate (Figure 1A and C below), typically when distance metrics as used in PCA cannot discriminate between samples that do not share any common features. Figure 1A below shows an isolation by distance or “band” pattern, in which neighboring samples have mostly similar alleles, but the similarity in allelic composition declines with the distance between samples. For example, “Sample 0” has the reference alleles at SNPs 1-10, but does not share the same alleles at any SNPs with samples 10-20. As a result, the euclidean distance (on which a PCA is based) saturates with increasing distance between samples (Figure 1C below).

@@

## Relationship between PCA and Admixture

>>>>

```{r setup, eval=TRUE, echo=TRUE, message = FALSE, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, cache = TRUE) 
```
Authors: KE Lotterhos, DL Davenport, S Truskey <br>
Acknowledgements/ Page Reviewers: Jason Johns

@@

On this page we outline and describe many aspects of PCA as it relates to population genomics:

* [What is PCA and what are its common uses in population genetics?](#whatspca)
* Considerations before sampling:
    + [How one chooses to sample from a landscape](#sampling)
    + [Sample sizes](#samplesize)
* Considerations while processing data (aka 'weird stuff that happens in PCA'): 
    + [Batch effects](#batcheffects)
    + [LD and inversion effects](#ld)
    + [Missing data and imputation](#missingdata)
    + [Hybridization and inbreeding effects](#hybrid)
    + [Horseshoe and wave patterns](#horseshoe)
* Comparison to other common approaches to investigating population genetic structure
    + [PCA  vs ADMIXTURE/STRUCTURE](#pca_admix)

@@

## What is PCA and what are its common uses in population genetics?{#whatspca}

@@

The eigenvectors are represented as a standard scatter plot, where points are samples/individuals represented on the new system of axes (PCs). Dimension reduction to eigenvectors (PCs) means that the 1st PC is the linear combination of the original variables that explains the most/greatest amount of variation, while the 2nd PC accounts for the greatest amount of the remaining variation being orthogonal (uncorrelated) to the 1st PC, the 3rd PC accounts for the greatest amount of the remaining variation being orthogonal (uncorrelated) to the 1st and 2nd PC so on and so on. In population genomics, typically the 1st PC will capture broad scale population structure (because broadly distributed populations are usually the most differentiated) while subsequent axes capture more regional population structure (usually less differentiated at smaller geographic scales, see Figure 2A). However, other cool examples using PCA find the greatest amount of variation is between individuals, particularly in the case of inversions, structural variation or regions associated with a specific phenotype or adaptation (see Figure 2B).

@@

* [Jolliffe, I. & Cadima, J. (2016) Principal component analysis: a review and recent developments](https://doi.org/10.1098/rsta.2015.0202)
* [Pearson K. (1901) On lines and planes of closest fit to systems of points in space. Philos Mag 2:559–572](https://doi.org/10.1080/14786440109462720)
* [Hotelling H (1936) Relations between two sets of variates. Biometrika 28:321–377](https://doi.org/10.1007/978-1-4612-4380-9_14)
* [Fentaw Abegaz et al. (2019) Principals about principal components in statistical genetics, Briefings in Bioinformatics](https://doi.org/10.1093/bib/bby08)
* [Josse J., Husson F. (2012) Handling missing values in exploratory multivariate data analysis methods. J. Soc. Française Stat., 153, 79–99.](http://www.numdam.org/item/JSFS_2012__153_2_79_0/)
* [François O, Gain C (2021) A spectral theory for Wright’s inbreeding coefficients and related quantities. PLoS Genet 17: e1009665](https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1009665)
* [Patterson N, Price AL, Reich D (2006) Population structure and eigenanalysis. PLoS Genet 2: e190](https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.0020190)
* [Bakker JD (2023) PCA. Applied Multivariate Statistics in R. University of Washington](https://uw.pressbooks.pub/appliedmultivariatestatistics/chapter/pca/)

@@

## Considerations before sampling
### How one chooses to sample from a landscape{#sampling}
How samples are selected from a metapopulation also affects the visual pattern in a PCA.

@@

### Sample sizes{#samplesize}
Uneven sampling between groups, the number of samples and/or the number of markers/SNPs used in PCA can change the PCA projection space, and thus interpretation of results, relative to the true demographic history of the sampled groups.

@@

Missing data in population genomic analysis is common.  Sometimes it is random, or it may be a feature of the data type (i.e. low coverage NGS, microsatellite null alleles, mutation at RAD cut sites). In general, if missing data reflects a true population signal then it requires careful consideration during analysis.

@@

Implementations of PCA in genomics typically (a) avoid missing values (i.e pcadapt, plink) usually by first determining the covariance-matrix or genomic relationship matrix (GRM)) between each pair of individuals using the variables (loci) that are available for both samples or by (b) imputing missing values. Imputation is the process of replacing a missing value with a numerical value via inference. Imputation of missing data can be done by sophisticated methods often seen in genomic association studies, or if you have linked haplotypes or strong genotype-phenotype correlations (see @Yi2022-bi). However, in population genomic studies for species of conservation concern which may have few genomic resources, simple imputations such as mean imputation (replace missing with column mean) is commonly encountered.

Imputation via the mean genotype in the metapopulation.
In this case, missing values for all individuals at a locus are replaced with the mean for that locus (also called a mean-imputed matrix). Importantly, when the mean-imputed matrix is centered and scaled, the original missing genotypes become non-informative in PCA (missing values are placed at the origin). This can be problematic if you have a lot of missing data in your dataset, especially if that missing data is biased (not random). @Yi2022-bi demonstrated this effect using different simulated populations (no migration. 5% migration, isolation-by-distance-cline) with 1%, 10% and 20% missing data, either randomly distributed among individuals, or biased among some individuals and some populations. They conducted PCA using the adegenet::glPca() implementation of PCA with default mean imputation and standardization, retaining all PCA axes. They found that individuals biased with missing data would be dragged away from their real population clusters to the origin of PCA plots, making them indistinguishable from known admixed individuals, potentially leading to misinterpreted population structure (Figure 11).

@@

In summary, the choice of imputation can have drastic effects on inference if there is a lot of missing data. When there is population structure and missing genotypes are imputed with the mean genotype across all populations, the genetic differentiation among populations may be artificially reduced and the metapopulation would appear to be panmictic.  When there is no population structure and missing genotypes are imputed with the mean genotype within each sampling location, the genetic differentiation among populations may be artificially inflated and it could appear that there is genetic structure when there is not.

Since most genomic data has missingness, it may be preferable to implement a method of PCA which explicitly accounts for it (i.e pcadapt, PCAngsd).

@@

### Horseshoe and Wave patterns{#horseshoe}
Horseshoe patterns arise in population genetic data that has isolation by distance structure, by which genetic differentiation among locations increases with the distance between them. The result when this type of data is put into a PCA is that the pattern looks like a horseshoe (Figure 1B from @Morton2017-xy below). In Figure 1B, Sample 0 is most distantly related from Sample 20, but they have the same loading on the PC1 axis. Horseshoe patterns also arise in microbial ecology, where they were originally misinterpreted as having an ecological explanation before they were proven to be a statistical artifact (@Morton2017-xy). The pattern arises in microbial data or genetic data as a consequence of distance metrics that saturate (Figure 1A and C below), typically when distance metrics as used in PCA cannot discriminate between samples that do not share any common features. Figure 1A below shows an isolation by distance or “band” pattern, in which neighboring samples have mostly similar alleles, but the similarity in allelic composition declines with the distance between samples. For example, “Sample 0” has the reference alleles at SNPs 1-10, but does not share the same alleles at any SNPs with samples 10-20. As a result, the euclidean distance (on which a PCA is based) saturates with increasing distance between samples (Figure 1C below).

@@

## Relationship between PCA and Admixture{#pca_admix}
